Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-10-15 fstats_pca_lewontin.tex
revision text updates

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
\end{abstract}
\section{Introduction}
As in other species, the genetic diversity of human populations  is shaped by their environment and history over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994, schraiber2015}. In turn, studying the diversity patterns in humans can be used to reconstruct the demographic and evolutionary history of our species. As such, population genetic theory is essential to understand how this works.

\paragraph{processes}
Over time, genetic drift will slowly change allele frequencies and as a result, isolated populations are expected to slowly diverge. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance. However, isolation-by-distance patterns are not uniform, but shaped by geography, particularly barriers   to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}. In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2016, racimo2020}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity.

Modern descriptions of human genetic diversity focus on the evolutionary processes that caused it, and which gives rise to  both discrete and continuous components \citep{rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a}.
The interplay of these demographic processes leads to the complex genetic structure that we observe in present-day people, and which we aim to reconstruct from genetic data, which is challenging since we cannot expect to devise a single model that captures all complexity.

\paragraph{Data-driven approaches}
One commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions; each emphasizing a particular aspects of the data. Particularly for large data sets with several thousand samples, exploratory, data-driven methods such as population trees, Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are a first step aimed at providing a simple, low-dimensional summary of the data while making minimal assumptions about the underlying population structure.

\paragraph{Model-based appraoches}
While PCA and related tools give a good overview of the data, they are not designed to answer specific research questions and cannot be used for model testing or parameter estimates. For this purpose, methods based on explicit demographic models  are often used \citep{gutenkunst2009, kamm2015, excoffier2013}. The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modelling assumptions such as that populations are discrete, randomly mating, and at equilibrium, with the hope that despite these assumptions being violated, the resulting model fits provide a sufficiently accurate description to answer the research questions.

\paragraph{F-stats}
However, when the number of populations exceeds a few dozen,  even codifying reasonable population models including all data can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset. However, this has the drawback that a large proportion of the data may be unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.

The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. I safe the formal definition for later; but the easiest way to motivate them assumes that populations are related as a tree, where the edges measure how much genetic drift has occurred. In this case, $F_2$ measures the genetic distance between populations, $F_3$ corresponds to an external, and $F_4$ to an internal branch of a tree ) \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

In most applications, these $F$-statistics are estimated from some data, and then used as tests of treeness. The canonical alternative model is an admixture graph or phylogenetic network \citep{patterson2012}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A). Typically, analyses proceed by testing specific hypotheses abot gene flow using $F$-statistics \citep{reich2009, green2010, lazaridis2014}. Increasingly commonly, $F$-statistics are also incorporated into tools that jointly fit more complicated models \citep[e.g. ][]{patterson2012,harney2021}. However, admixture graphs are not the only alternative model that explain violations of treeness. Expected $F$-statistics can be calculated for a wide range of demographic models \citep{peter2016}, or even in abstract Euclidean spaces \citep{oteo-garcia2021}, so that model fit remains a major concern.

\paragraph{PCA}
With minimal assumptions, PCA is one of the most widely used data summary and data exploration techniques. One way to think about PCA is that it aims to create a low-dimensional summary of the data that retains a maximum of information about the underlying structure in the data (Again, a formal description will be given in the next section). It has been widely used and independently discovered in many disciplines. In population genetics, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize global genetic diversity in the pre-genomic era \citep{cavalli-sforza1994}. Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale population genetic data set.  This is most commonly done in early, exploratory phases of a study \citep{schraiber2015}, as PCA can be used to e.g. remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses. Theoretically, the PCA-decomposition has been studied in terms of trees \citep{cavalli-sforza1975}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}.

\paragraph{F-stats and PCA}
However, the connections between PCA and $F$-statistics remains poorly understood, and the goal of this paper is to investigate this connection. To do so, I use the geometric interpretation of $F$-statistics developed by \cite{oteo-garcia2021} who defined them in arbitrary Euclidean spaces. We make use of the empirical finding that while the allele-frequency space is frequently high-dimensional, for many empirical data sets, a low number of dimensions are sufficient to characterize population structure, and hence $F$-statistics. This allows for approximate interpretations of $F$-statistics on PCA plots, and I use two sample data sets to evaluate these approximations in practice.

\section{Theory}
In this section, I will introduce the mathematics and notations for $F$-statistics and PCA. A comprehensive  treatise on PCA is given by e.g. \cite{jolliffe2013} and \cite{pachter2014}, and a useful guide to interpretation is \cite{cavalli-sforza1994}. Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.

\subsection{Formal Definition to $F$-statistics}
Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ loci. Let $x_{il}$ denote the frequency at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$. As $X_i$ will be the only data summary for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.

@@

As highlighted in the introduction, $F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}. In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of all branches between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch between four nodes, respectively (Figure \ref{fig:geom}C). Crucially, for branches that do not exist in the tree (as in Figure \ref{fig:geom}D), $F_4$ will be zero. The length of each branch can be thought of in units of genetic drift, and is non-negative \citep{patterson2012}.

@@

Most commonly however, $F_3$ and $F_4$ are used as tests of treeness \citep{patterson2012}: Negative $F_3$-values correspond to a branch with negative genetic drift, which is a violation of treeness. Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{patterson2012}. The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A, where (the typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the population we observe. This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact equations can be found in ][]{peter2016}. While admixture graphs are commonly considered, other alternative models exist. In \citep{peter2016}, I showed that $F_4$ is sensitive to many non-symmetric population structure models. $F_3$ is more robust, but may also be negative if substantial population substructure exists.

\subsubsection{Geometric interpretation of $F$-statistics}
For very large data sets, it is hard to justify a specific demographic models that includes all data, but we will typically still be able to explore such data e.g. using PCA. One way to justify $F$-statistics is to interpret the populations $X_i$ geometrically as points or vectors in the $S$-dimensionsl \emph{data space} $\mathbb{R}^S$.  \cite{oteo-garcia2021} showed that in this case, the $F$-statistics can be thought of as inner (or dot) products, and that all properties and tests related to treeness can be derived from this larger space. In particular the $F$-statistics can be defined as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. \cite{oteo-garcia2021} showed that in this framework, $F$-statistics are closely related to vector projections
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$.

The drawback of the geometric approach of \cite{oteo-garcia2021} is that we have to deal with an very high-dimensional space, as the number of SNPs is frequently in the millions. However, it has been commonly observed that population structure is quite low-dimensional, with the first few PCs providing a good approximation of the underlying population structure. Therefore, we may hope that PCA could yield a reasonable approximation of the data space.

@@

\subsection{Formal Definition of PCA}
PCA is a common way of summarizing genetic data, and so a large number of variation of PCA exist, e.g. in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis. The version of PCA I describe here is set up in a way that the similarities to the $F$-statistics-framework are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets. In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}. The differences this causes will be addressed in the discussion.

Let us again assume we have genotype data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.

One way one can define PCA is that it aims to find a $K$-dimensional subspace of the data space that retains the most important variation. $K$ is at most $n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

	where $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$  the identity matrix and a matrix of ones, respectively. For any matrix $\MY$, we can perform a singular value decomposition $\MY = \mathbf{U}\MSINGULAR\mathbf{V}^T$ which, in the context of PCA, is interpreted as follows: The matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is contains information about population structure. The SNP loadings $\ML=\mathbf{V}^T$ form an orthonormal basis of size $n \times S$, its rows give the contribution of each SNP to each PC. It is often used to look for outliers, which might be indicative of selection \citep[e.g][]{duforet-frebourg2016}. Alternatively, the PCs can also be motivated as an eigendecomposition of the covariance matrix $\MY\MY^T$. This can be motivated from (\ref{eq:svd}):

@@

\subsection{Connection between PCA and $F$-statsitics}	
\subsubsection{Principal components from $F$-statistics}
PCA and $F$-statistics are closely related. In fact, the principal components can be directly calculated from $F$-statistics using multidimensional scaling, which, for squared Euclidean ($F_2$)-distances, leads to an identical decomposition to PCA \citep{gower1966}. Suppose we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$. We can obtain the principal components from this matrix by double-centering it, so that its row and column means are zero, and perform an eigendecomposition of the resulting matrix:
\begin{equation}
\MP\MP^T = - \frac{1}{2}\MC\MF\MC \text{.} \label{eq:mds}
\end{equation}

\subsubsection{$F$-statistics in PCA-space}
By performing a PCA, we rotate our data to reveal the axes of highest variation. However, the dot product is invariant under rotation, and $F$-statistics can be thought of as dot products.  What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or any other orthogonal basis such as the principal components $\MP$. Formally,

@@

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{align}
F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
\end{align}
In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error. I will omit the superscript when the exact number of PCs is not relevant.
If we sum up the squared approximation errors over all pairs of populations, we obtain 
\begin{equation}
\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
\end{equation}
where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements. This is precisely the function that is minimized in MDS \citep{jolliffe2013}. In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the sum of approximation errors of all $F_2$-statistics.

\subsubsection{$F$-statistics and projection on PCA}
One of the easiest ways of dealing with missing data in PCA is to calculate the principal components (equation \ref{eq:svd}) only on a subset of the data with no missingness, and then to \emph{project} the lower quality samples with high missingness onto this PCA. The simplest way to do this is to note that

>>>>

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
\end{abstract}
\section{Introduction}
As most species, the genetic diversity of human populations is influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, nielsen2020}. Population genetic models use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

In particular, in isolated populations genetic drift will slowly change allele frequencies. As a result, isolated populations are expected to slowly differentiate \citep{wahlund1928,cavalli-sforza1975}. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985a}. However, isolation-by-distance patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}. In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016} and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{skov2019}.

The interplay of these demographic processes leads to the complex genetic structure observed in present-day human populations \citep{the2015, mallick2016} with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a} that we model in order to reconstruct human demographic history. This is challenging because particularly for large and heterogeneous data sets we cannot expect to devise a single model that captures all processes. A commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions. each emphasizing particular aspects of the data.

A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}. Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}). However, these methods are not designed to answer specific research questions, and are limited in their ability to estimate biologically meaningful parameters. For this purpose, methods based on explicit demographic models are often used that aim to fit a specified or estimated model of divergence, migration and genetic drift to the data \citep{gutenkunst2009, excoffier2013, kamm2015}. The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modeling assumptions such as that populations are discrete, randomly mating, or at equilibrium. While in most cases these assumptions are violated to some extent and cannot be verified, but we hope that the resulting model fits provide a sufficiently accurate answer for specific research questions.

\paragraph{F-stats}
However, when the number of populations exceeds a few dozen,  even codifying reasonable population models can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}. However, this has the drawback that a large proportion of the data may be unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.

The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. I save the formal definition for later; but the easiest way to motivate them assumes that populations are related as a tree, where the edge lengths measure how much genetic drift has occurred. \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

In most applications, these $F$-statistics are estimated from  data, and then used as tests of treeness. In particular, under the assumption of a tree, $F_3$ is restricted to be non-negative, and many $F_4$-statistics will be zero \citep{semple2003, patterson2012}, and data that violates these constraints is incompatible with a tree-like relationship between populations. The canonical alternative model is an admixture graph (or phylogenetic network) \citep{patterson2012, huson2010}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A). However, admixture graphs are not the only plausible alternative model, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.

\paragraph{F-stats and PCA}
The practical issue addressed in this study is how $F$-statistics can be reconciled with one of the most widely used data-driven  techniques, PCA. One way PCA can be motivated is as generating a low-dimensional representation of the data, with each dimension (called principal components, PCs) retaining a maximum of the variance present in the data. In population genetics, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize global genetic diversity \citep{cavalli-sforza1994}. Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale data sets. 
%This is most commonly done in early, exploratory phases of a study \citep{schraiber2015}, as PCA can be used to e.g. remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses. 
The PCA-decomposition has been studied for a number of population genetic models including trees \citep{cavalli-sforza1975}, spatially continuous structure \citep{novembre2008a}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}. Here, in order to link PCA to $F$-statistics, I interpret both of them geometrically in \emph{allele frequency space}, i.e. as functions of a high-dimensional Euclidean space. For $F$-statistics, this interpretation was recently developed by \cite{oteo-garcia2021}, and for PCA it follows naturally from the interpretation of approximating a high-dimensional space with a low-dimensional one.

In the next section, I will formally derive the connection between $F$-statistics and PCA, and show how $F$-statistics can be interpreted geometrically, with a particular emphasis on two-dimensional PCA plots. In the results section, I will then discuss how some of the most common applications of $F$-statistics manifest themselves on a PCA, and illustrate them on two example data sets.

\section{Theory}
In this section, I will introduce the mathematics and notations for $F$-statistics and PCA. A comprehensive  treatise on PCA is given by e.g. \cite{jolliffe2013} a useful primer on the mathematics is \cite{pachter2014}, and a useful guide to interpretation is \cite{cavalli-sforza1994}. Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.

\subsection{Formal Definition of $F$-statistics}
Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ loci. Let $x_{il}$ denote the frequency at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$. As $X_i$ will be the only data summary considered here for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.

@@

As highlighted in the introduction, $F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}. In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of the path between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch between four nodes, respectively (Figure \ref{fig:geom}C). Crucially, for branches that do not exist in the tree (as in Figure \ref{fig:geom}D), $F_4$ will be zero. The length of each branch can be thought of in units of genetic drift, and is non-negative \citep{patterson2012}.

@@

Most commonly however, $F_3$ and $F_4$ are used as tests of treeness \citep{patterson2012}: Negative $F_3$-values correspond to a branch with negative genetic drift, which is a violation of treeness. Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{patterson2012}. The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A, where (the typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the population we observe. This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in ][]{peter2016}.

\subsubsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Recently, \cite{oteo-garcia2021} showed that these assumptions are not necessary by re-deriving $F$-statistics in a geometric framework. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and that all properties and tests related to treeness can be derived from this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to vector projections
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$.

The drawback of the geometric approach of \cite{oteo-garcia2021} is that we have to deal with an very high-dimensional space, as the number of SNPs is frequently in the millions. However, it has been commonly observed that population structure is quite low-dimensional, and that the first few PCs provide a good approximation of the covariance structure in the data \citep{patterson2006}. Therefore, we may hope that PCA could yield a reasonable approximation of the allele frequency space, and that $F$-statistics as measures of population structure may likewise be well-approximated by the first few PCs.

@@

\subsection{Formal Definition of PCA}
PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g. in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis. The version of PCA I describe here is set up in a way that the similarities to the $F$-statistics framework are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets. In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}. The differences this causes will be addressed in the discussion.

Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.

One way PCA can be motivated is that it aims to find a $K$-dimensional subspace of the data space that retains most  variation in the data. $K$ is at most $n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

	where $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$  the identity matrix and a matrix of ones, respectively. For any matrix $\MY$, we can perform a singular value decomposition $\MY = \mathbf{U}\MSINGULAR\mathbf{V}^T$ which, in the context of PCA, is interpreted as follows: The matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and contains information about population structure. The SNP loadings $\ML=\mathbf{V}^T$ form an orthonormal basis of size $n \times S$, its rows give the contribution of each SNP to each PC. It is often used to look for outliers, which might be indicative of selection \citep[e.g][]{duforet-frebourg2016}. Alternatively, the PCs can also be obtained from an eigendecomposition of the covariance matrix $\MY\MY^T$. This can be motivated from (\ref{eq:svd}):

@@

\subsection{Connection between PCA and $F$-statsitics}	
\subsubsection{Principal components from $F$-statistics}
PCA, as defined above, and $F$-statistics are closely related. In fact, the principal components can be directly calculated from $F$-statistics using multidimensional scaling, which, for squared Euclidean ($F_2$)-distances, leads to an identical decomposition to PCA \citep{gower1966}. Suppose we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$. We can obtain the principal components from this matrix by double-centering it, so that its row and column means are zero, and perform an eigendecomposition of the resulting matrix:
\begin{equation}
\MP\MP^T = - \frac{1}{2}\MC\MF\MC \text{.} \label{eq:mds}
\end{equation}

\subsubsection{$F$-statistics in PCA-space}
By performing a PCA, we rotate our data to reveal the axes of highest variation. However, the dot product is invariant under rotation, and $F$-statistics can be thought of as dot products \citep{oteo-garcia2021}.  What this means is that we are free to calculate $F_2$ either on the uncentered data $\MX$, the centered data $\MY$ or any other orthogonal basis such as the principal components $\MP$. Formally,

@@

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{align}
F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
\end{align}
In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error. I will omit the superscript of $\hat{F_2}$ when the exact number of PCs is not relevant.
If we sum up the squared approximation errors over all pairs of populations, we obtain 
\begin{equation}
\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
\end{equation}
where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements. This is precisely the function that is minimized in MDS \citep{jolliffe2013}. In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the sum of approximation errors of all $F_2$-statistics.

\subsubsection{$F$-statistics and samples projected onto PCA}
One of the easiest ways of dealing with missing data in PCA is to calculate the principal components (equation \ref{eq:svd}) only on a subset of the data with no missingness, and then to \emph{project} the lower quality samples with high missingness onto this PCA. The simplest way to do this is to note that

\begin{equation*}
\MY\ML^T = \MP\ML\ML^T = \MP,
\end{equation*}
and so a new (centered) population $Y_{\text{new}}$ can be projected onto an existing PCA simply by post-multiplying it with $\ML^T$: $$P_{\text{proj}} = Y_{\text{new}}\ML^T;$$ the $k$-th entry of $P_{\text{proj}}$ gives the coordinates of the new sample on the $k$-th PC. However, it is likely that $Y_{\text{new}}$ lies outside the variation of the original sample. In this case, there is a projection error 
$$\normsq{Y_{\text{new}} - P_{\text{proj}}\ML } =  F_2(P_{\text{proj}}\ML, Y_{\text{new}}).$$ If we project with missing data, a similar projection can be used where we remove the rows from $Y_{\text{new}}$ and $\ML$ where data in $Y_{\text{new}}$ is missing \citep{patterson2006}.
